### 2024-09-18 Paper/MSEPM-Manuscript.tex

\abstract{Epigenetic clocks, DNA methylation based predictive models of chronological age, are often utilized to study aging associated biology. 
Despite their widespread use, these methods do not account for other factors that also contribute to the variability of DNA methylation data.
 For example, many CpG sites show strong sex-specific or cell type specific patterns that likely impact the predictions of epigenetic age.  
 To overcome these limitations, we developed a multidimensional extension of the Epigenetic Pacemaker, the Multi-State Epigenetic Pacemaker (MSEPM). 
 We show that the MSEPM is capable of accurately modeling multiple methylation associated factors simultaneously, while also providing site specific 
 models that describe the per site relationship between methylation and these factors. We utilized the MSEPM with a large aggregate cohort of blood
  methylation data to construct models of the effects of age, sex and cell type heterogeneity on DNA methylation.  We found that these models 
  capture a large faction of the variability at thousands of DNA methylation sites.  Moreover, this approach allows us to identify sites that 
  are primarily affected by aging and no other factors.  An analysis of these sites reveals that those that lose methylation over time are 
  enriched for CTCF transcription factor chip peaks, while those that gain methylation over time are associated with bivalent promoters of 
  genes that are not expressed in blood. These observation suggest mechanisms that underlie age associated methylation changes and suggest 
  that age associated increases in methylation may not have strong functional consequences on cell states. In conclusion, the MSEPM is 
  capable of accurately modeling multiple methylation associated factors and the models produced can illuminate site specific 
  combinations of factors that affect methylation dynamics.}

@@

We implemented a simulation framework using the MSEPM formulation evaluate the performance of the MSEPM model under various conditions.
To simulate the association between methylation status and an observable factor we modeled the 
epigenetic factor $p_{k,j}$ as function of time, or age, and magnitude, $p_{k,j}^0$ with a non-linear transformation $\gamma_{k}$, 
where $p_{k,j} = Age_j p_{k,j}^{0,\gamma_{k}}$. In practice the value of the $p_{k,j}$ is often unknown and the association between 
methylation status and $p_{k,j}$ is inferred through the observable factor $p_{k,j}^0$.

@@

\subsection{H3K4me3 enrichment analysis} 
Enrichment of analysis for H3K4me3 (figure 7A) was carried out by downloading rpm normalized bigwig files of H3K4me3 ChIP-seq data from 
epigenomesportal\cite{Bujold2016-vk} for CD38+ B Cells and CD56+ NTK Cells (for both 0-5 years old and 60-65 years old individuals). 
Heatmaps of H3K4me3 were generated using deepTools2\cite{Ramirez2016-xl} using the computeMatrix and plotHeatmap function to plot the 
bigwig signal over genomic regions of cluster 7 as the BED input. The IGV genome browser\cite{Robinson2011-he} was used to generate an image of the KCTD1 
and IRS2 promoter regions shown in figure 7B using downloaded bigwig tracks.

@@

We simulated 90 methylation sites (Figure 1I). We then evaluated the MSEPM model as follows. We simulated 1000 samples with the 
four epigenetic factors described above. We then simulated methylation values using the simulated site rates. Simulated samples
 were then split for training ($n=500$) and testing ($n=500$). MSEPM models were then fitted using the values of the input factors, $p^0_{k,j}$. 
 We generated 1000 simulated datasets and fit MSEPM models using four combinations of input factors (Age, Age-Sex, Age-Sex-CN, Age-Sex-CN-CE). 
 Within each simulation, epigenetic state predictions and methylation site predictions were made for all testing samples. All models captured 
 the nonlinear association between simulated age and methylation (Supp. Figure 1). As the number of factors in the model is increased the mean 
 absolute error (MAE) between the predicted epigenetic states and the simulated epigenetic factors decreases (Figure 2A). Importantly, to accurately
  assess simulated age it is necessary to account for the influence of the other simulated factors (Sex, CN, CE). The MAE between the predicted 
  and simulated methylation values decreases as simulated factors are added to the model, and accurately assessing the methylation status of a 
  simulated site requires that the factor associated with the methylation status at the site is included in the model (Figure 2A).

>>>>

\abstract{Epigenetic clocks, DNA methylation based predictive models of chronological age, are often utilized to study aging associated biology. 
Despite their widespread use, these methods do not account for other factors that also contribute to the variability of DNA methylation data.
 For example, many CpG sites show strong sex-specific or cell type specific patterns that likely impact the predictions of epigenetic age.  
 To overcome these limitations, we developed a multidimensional extension of the Epigenetic Pacemaker, the Multi-State Epigenetic Pacemaker (MSEPM). 
 We show that the MSEPM is capable of accurately modeling multiple methylation associated factors simultaneously, while also providing site specific 
 models that describe the per site relationship between methylation and these factors. We utilized the MSEPM with a large aggregate cohort of blood
  methylation data to construct models of the effects of age, sex and cell type heterogeneity on DNA methylation.  We found that these models 
  capture a large faction of the variability at thousands of DNA methylation sites.  Moreover, this approach allows us to identify sites that 
  are primarily affected by aging and no other factors.  An analysis of these sites reveals that those that lose methylation over time are 
  enriched for CTCF transcription factor chip peaks, while those that gain methylation over time are associated with bivalent promoters of 
  genes that are not expressed in blood. These observations suggest mechanisms that underlie age associated methylation changes and suggest 
  that age associated increases in methylation may not have strong functional consequences on cell states. In conclusion, the MSEPM is 
  capable of accurately modeling multiple methylation associated factors and the models produced can illuminate site specific 
  combinations of factors that affect methylation dynamics.}

@@

We implemented a simulation framework using the MSEPM formulation to evaluate the performance of the MSEPM model under various conditions.
To simulate the association between methylation status and an observable factor we modeled the 
epigenetic factor $p_{k,j}$ as function of time, or age, and magnitude, $p_{k,j}^0$ with a non-linear transformation $\gamma_{k}$, 
where $p_{k,j} = Age_j p_{k,j}^{0,\gamma_{k}}$. In practice the value of the $p_{k,j}$ is often unknown and the association between 
methylation status and $p_{k,j}$ is inferred through the observable factor $p_{k,j}^0$.

@@

\subsection{H3K4me3 enrichment analysis} 
Enrichment of analysis for H3K4me3 (Figure 7A) was carried out by downloading rpm normalized bigwig files of H3K4me3 ChIP-seq data from 
epigenomesportal\cite{Bujold2016-vk} for CD38+ B Cells and CD56+ NTK Cells (for both 0-5 years old and 60-65 years old individuals). 
Heatmaps of H3K4me3 were generated using deepTools2\cite{Ramirez2016-xl} using the computeMatrix and plotHeatmap function to plot the 
bigwig signal over genomic regions of cluster 7 as the BED input. The IGV genome browser\cite{Robinson2011-he} was used to generate an image of the KCTD1 
and IRS2 promoter regions shown in Figure 7B using downloaded bigwig tracks.

@@

We simulated 90 methylation sites (Figure 1I). We then evaluated the MSEPM model as follows. We simulated 1000 samples with the 
four epigenetic factors described above. We then simulated methylation values using the simulated site rates. Simulated samples
 were then split for training ($n=500$) and testing ($n=500$). MSEPM models were then fitted using the values of the input factors, $p^0_{k,j}$. 
 We generated 1000 simulated datasets and fit MSEPM models using four combinations of input factors (Age, Age-Sex, Age-Sex-CN, Age-Sex-CN-CE). 
 Within each simulation, epigenetic state predictions and methylation site predictions were made for all testing samples. All models captured 
 the nonlinear association between simulated age and methylation (Supp. Figure 1). As the number of factors in the model is increased the mean 
 absolute error (MAE) between the predicted epigenetic states and the simulated epigenetic factors decreases (Figure 2A). Importantly, to accurately
  assess simulated age it is necessary to account for the influence of the other simulated factors (Sex, CN, CE).
